---
title: "Data Manipulation in R (BIO294)"
author: "Nikos Minadakis"
date: "10/11/2021"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Documents/Project/Bio294_teaching/2021/2_Data_Manipulation_R/2_scripts/')
```

# **Getting started**

----

clean the environment
```{r message=FALSE}
rm(list = ls())
```

set working directory
```{r message=FALSE, eval=FALSE}
setwd()
```

libraries are sets of functions that expand the possibilities of base R
```{r message=FALSE}
library(tidyverse)
```

calculator
```{r}
1 + 3
2 * 7
```

input data (arrows are assigning values to objects)
```{r}
Bdis <- read.csv('../1_data/Bdis_matrix.csv')
```

get help
```{r}
?read.csv
```


**the data set**

# **Data preparation**

----

## *Data exploration*

functions are always working on what is inside brackets.
options are also defined there

```{r}
str(Bdis)
```

```{r}
head(Bdis)
```

first row
```{r}
Bdis[1,]
```

first column (combined with head())
```{r}
head(Bdis[,1])
```

length of vector
```{r}
length(Bdis[,1])
```

combined
```{r}
Bdis[1,1]
```

the highest value for only one column
```{r}
max(Bdis$longitude)
```

what is the position of the highest value
```{r}
which.max(Bdis$longitude)
```

```{r}
summary(Bdis)
```

### **as_tibble**

```{r}
Bdis.t <- as_tibble(Bdis)
```

```{r}
Bdis.t
```

# **Data manipulation**

----

**dplyr**

- filter
- select
- mutate
- group_by
- summarise

**tidyr**

- pivot_longer

## *filter*

filter rows
```{r}
filter(Bdis.t, cluster == "C")
```

### **%>% and %in%**

- %>%(pipe): the output is used as input for the next function
- %in%: this operator can be used to identify if an element belongs to a vector or dataframe
```{r}
Bdis.t %>%
  filter(cluster == "C" & source == "ar_cs")
```

```{r}
Bdis.t %>%
  filter(cluster %in% c("C", "A_Italia"))
```

```{r}
my_clusters <- c("C", "A_Italia")
Bdis.t %>%
  filter(cluster %in% my_clusters)
```

## *select*

select columns
```{r}
Bdis.t %>%
  select(names)
```

```{r}
Bdis.t %>%
  select(names, cluster)
```

### **starts_with, ends_with, contains**

```{r}
Bdis.t %>%
  select(starts_with("B"))
```

```{r}
Bdis.t %>%
  select(ends_with("East"))
```

```{r}
Bdis.t %>%
  select(contains("names"))
```

```{r}
Bdis.t %>%
  select(contains("names"), cluster)
```

## *mutate*

```{r}
Bdis.t$names <- as.character(Bdis.t$names)
Bdis.t$altern_names <- as.character(Bdis.t$altern_names)
```

### **ifelse** and **is.na**

- ifelse('question', 'what happens if TRUE', 'what happens if FALSE')
- is.na gives TRUE if the value is NA, and FALSE if not
- mutate creates a new value in the data frame ('new_names')
```{r}
Bdis.t %>% 
  mutate(new_names = ifelse(is.na(Bdis.t$altern_names), Bdis.t$names, Bdis.t$altern_names))
```

## *group_by* and *summarise*

- group_by takes an existing table and converts it into a grouped table where operations are performed "by group".
Grouping doesn't change how the data looks, it changes how it acts with the other dplyr verbs
- summarise creates a new data_frame that has one row for each group variable
```{r}
Bdis.t %>%
  group_by(cluster) %>%
  summarise(max_lon = max(longitude))
```


n = n() creates a column in summary with a function (n) that writes the number
```{r}
Bdis.t %>%
  group_by(cluster) %>%
  summarise(max_lon = max(longitude),
            min_lon = min(longitude),
            mean_lon = mean(longitude),
            n = n())
```

## *pivot_longer*

pivot_longer lengthens data, increasing the number of rows and decreasing the number of columns
```{r}
Bdis_plot <- Bdis.t %>%
  pivot_longer(names_to = "clade", cols = C:B_East, values_to = "percent")
```

### **visualize**

choose one individual
```{r}
Bdis_plot <- Bdis.t %>%
  pivot_longer(names_to = "clade", cols = C:B_East, values_to = "percent") %>%
  filter(names == "Cb23")

as_tibble(Bdis_plot)
```

color palette
```{r}
library(rcartocolor)
my_palette_k5 <- c("violetred4",  "#21908CFF", "#75D054FF","#404688FF", "#FDE725FF")
```

bar plot and then pie chart
```{r}
Bdis_plot <- Bdis_plot %>%
  ggplot(aes(x = "", y = percent, fill = clade)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0)
```

make it pretty
```{r}
Bdis_plot <- Bdis_plot +
  scale_fill_manual(values = my_palette_k5) +
  labs(x = NULL, y = NULL, title = "Cb23 ancestry proportions") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

visualize
```{r}
Bdis_plot
```